Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Tumors

### 2018-06-04 submission/manuscript.Rmd
smooth over the results

**Decreased SCFA concentrations are not associated with tumors.** We used frozen fecal samples from `r length(metai$sample)` individuals and HPLC to measure acetate, butyrate, and propionate concentrations at a cross-sectional point in time. There was no difference between individuals with normal colons (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F]. Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.

**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models to classify normal versus adenoma and normal versus carcinoma with OTUs or OTUs and SCFA concentrations. For both adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although including SCFA concentrations did not add extra information for classification of disease using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may change due to disease status; where a smaller number of microbes are responsible for the observed SCFA concentrations.

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and performed metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to tumors. Although these results suggest that SCFA concentrations do not change due to tumor, there may be some errors present within our results that may account for the lack of associations.

**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A]. However, these models tended to be overfit, suggesting that rarer taxa may be important for this classification [Figure S1A]. The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B]. These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate. Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations. Overall, our results do not support the hypothesis that lower SCFA concentrations are associated with colorectal cancer.

>>>>

**Decreased SCFA concentrations are not associated with tumors.** We used high-performance liquid chromatography (HPLC) to measure acetate, butyrate, and propionate concentrations of frozen fecal samples from `r length(metai$sample)` individuals at a cross-sectional point in time. There was no difference between individuals with normal colons (n=`r length(filter(metai, dx == "normal") %>% pull(sample))`) and patients with either an adenoma (n=`r length(filter(metai, dx == "adenoma") %>% pull(sample))`) or carcinoma (n=`r length(filter(metai, dx == "cancer") %>% pull(sample))`) for any of the SCFAs measured after multiple comparison correction (P-value `r ifelse(min(pull(hplc_groups_scfa_kruskal, bh)) > 0.15, invisible(">"), invisible("<"))` 0.15) [Figure 1A - 1C]. We next measured the concentration of SCFAs in `r length(metaf$initial)` patients with an adenoma (n=`r length(filter(metaf, dx == "adenoma") %>% pull(initial))`) or carcinoma (n=`r length(filter(metaf, dx == "cancer") %>% pull(initial))`) in which we had pre- and post-treatment fecal samples. Although there was a general trend for increasing acetate, butyrate, and propionate concentrations after treatment for tumors, there was no significant difference pre- and post-treatment for patients with adenoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "adenoma") %>% pull(bh)) > 0.20, invisible(">"), invisible("<"))` 0.20) or carcinoma (P-value `r ifelse(min(filter(hplc_treat_scfa_kruskal, disease == "carcinoma") %>% pull(bh)) > 0.80, invisible(">"), invisible("<"))` 0.80) [Figure 1D - 1F]. Even though there was no change in SCFA concentrations between individuals with normal colons and those with tumors, this information could still be important to help classify disease.

**Random Forest models with SCFA concentrations do not classify tumors better.** Using the Random Forest algorithm we built models with OTUs only or OTUs and SCFA concentrations to classify normal versus adenoma and normal versus carcinoma fecal samples. For adenoma and carcinoma models, there was no difference between the median AUC of models with or without SCFA concentrations (P-value `r ifelse(min(pull(disease_rf, bh)) > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 2]. Although including SCFA concentrations did not add extra information for classification of colorectal tumors using Random Forest models, it is possible that the genes for enzymes involved in SCFA synthesis may vary based on the type of colorectal tumor; where a smaller number of microbes are responsible for the observed SCFA concentrations in patients with carcinoma.

**Changes in genes for enzymes involved in SCFA synthesis are not associated with tumors.** Using a list of specific genes that are important for the synthesis of SCFAs [Table S1], we looked for differences in gene abundance based on individuals having normal colons, adenomas, or carcinomas. First, using imputed gene relative abundance based on 16S rRNA gene sequencing we found no difference in any of the genes inolved with acetate, butyrate, or propionate synthesis (P-value `r ifelse(min(pull(picrust_select, bh)) > 0.90, invisible(">"), invisible("<"))` 0.90) [Table S2]. This similarity between groups is highlighted by visualizing genes important in butyrate synthesis [Figure 3A]. Using a Wilcoxson rank-sum test, there also was no difference in imputed gene relative abundance between pre- and post-treatment samples for any genes involved with SCFA synthesis (P-value `r ifelse(min(pull(picrust_select_treat, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S3]. Next, we took a subset of these `r length(metai$sample)` fecal samples (n=`r length(geof_samples$X1)`) and used metagenomic sequencing to confirm these results. Like the imputed gene results, metagenomic analysis found that there was no difference in any of the genes involved in SCFA synthesis between individuals with normal colons (n=`r length(filter(geof_samples, X30 == "Healthy") %>% pull(X2))`) and patients with adenoma (n=`r length(filter(geof_samples, X30 == "Adenoma") %>% pull(X2))`) or carcinoma (n=`r length(filter(geof_samples, X30 == "Cancer") %>% pull(X2))`) (P-value `r ifelse(min(pull(opf_select, bh)) > 0.70, invisible(">"), invisible("<"))` 0.70) [Table S4]. This lack of difference is highlighted when we visualize the results for butyrate kinase [Figure 3B]. These observations provide evidence that the gene content does not change due to colorectal tumors. Although these results suggest that SCFA concentrations do not change due to colorectal tumor, there may be some errors present within our results that may account for the lack of associations.

**Expected taxa are associated with higher SCFA concentrations regardless of tumor status.** Using OTU data we built Random Forest models to classify higher than median and lower than median SCFA concentrations. Overall, OTU data had a reasonable ability to classify high and low SCFA concentrations [Figure S1A]. However, these models tended to be overfit, suggesting that rarer taxa may be important for this classification [Figure S1A]. The most important OTUs to these models (assessed with mean decrease in accuracy (MDA)) were to taxa that are normally associated with SCFA production [Figure S1B]. These results highlight that there are no immediately apparent errors in our data because SCFA concentrations are associated with taxa known to produce acetate, butyrate, and propionate. Additionally, OTUs associated with these taxa are the most important to models that can classify high and low SCFA concentrations. Overall, our results are robust and do not support the hypothesis that differences in SCFA concentrations are associated with colorectal cancer.

### 2018-06-04 submission/manuscript.tex
smooth over the results

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used frozen fecal samples from 490 individuals and HPLC to measure
acetate, butyrate, and propionate concentrations at a cross-sectional
point in time. There was no difference between individuals with normal
colons (n=172) and patients with either an adenoma (n=198) or carcinoma
(n=120) for any of the SCFAs measured after multiple comparison
correction (P-value \textgreater{} 0.15) {[}Figure 1A - 1C{]}. We next
measured the concentration of SCFAs in 67 patients with an adenoma
(n=41) or carcinoma (n=26) in which we had pre- and post-treatment fecal
samples. Although there was a general trend for increasing acetate,
butyrate, and propionate concentrations after treatment for tumors,
there was no significant difference pre- and post-treatment for patients
with adenoma (P-value \textgreater{} 0.20) or carcinoma (P-value
\textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even though there was no
change in SCFA concentrations between individuals with normal colons and
those with tumors, this information could still be important to help
classify disease.

\textbf{Random Forest models with SCFA concentrations do not classify
tumors better.} Using the Random Forest algorithm we built models to
classify normal versus adenoma and normal versus carcinoma with OTUs or
OTUs and SCFA concentrations. For both adenoma and carcinoma models,
there was no difference between the median AUC of models with or without
SCFA concentrations (P-value \textgreater{} 0.05) {[}Figure 2{]}.
Although including SCFA concentrations did not add extra information for
classification of disease using Random Forest models, it is possible
that the genes for enzymes involved in SCFA synthesis may change due to
disease status; where a smaller number of microbes are responsible for
the observed SCFA concentrations.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Using a Wilcoxson rank-sum test, there also was no
difference in imputed gene relative abundance between pre- and
post-treatment samples for any genes involved with SCFA synthesis
(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
these 490 fecal samples (n=85) and performed metagenomic sequencing to
confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals with normal colons (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 3B{]}. These
observations provide evidence that the gene content does not change due
to tumors. Although these results suggest that SCFA concentrations do
not change due to tumor, there may be some errors present within our
results that may account for the lack of associations.

>>>>

\textbf{Decreased SCFA concentrations are not associated with tumors.}
We used high-performance liquid chromatography (HPLC) to measure
acetate, butyrate, and propionate concentrations of frozen fecal samples
from 490 individuals at a cross-sectional point in time. There was no
difference between individuals with normal colons (n=172) and patients
with either an adenoma (n=198) or carcinoma (n=120) for any of the SCFAs
measured after multiple comparison correction (P-value \textgreater{}
0.15) {[}Figure 1A - 1C{]}. We next measured the concentration of SCFAs
in 67 patients with an adenoma (n=41) or carcinoma (n=26) in which we
had pre- and post-treatment fecal samples. Although there was a general
trend for increasing acetate, butyrate, and propionate concentrations
after treatment for tumors, there was no significant difference pre- and
post-treatment for patients with adenoma (P-value \textgreater{} 0.20)
or carcinoma (P-value \textgreater{} 0.80) {[}Figure 1D - 1F{]}. Even
though there was no change in SCFA concentrations between individuals
with normal colons and those with tumors, this information could still
be important to help classify disease.

\textbf{Random Forest models with SCFA concentrations do not classify
tumors better.} Using the Random Forest algorithm we built models with
OTUs only or OTUs and SCFA concentrations to classify normal versus
adenoma and normal versus carcinoma fecal samples. For adenoma and
carcinoma models, there was no difference between the median AUC of
models with or without SCFA concentrations (P-value \textgreater{} 0.05)
{[}Figure 2{]}. Although including SCFA concentrations did not add extra
information for classification of colorectal tumors using Random Forest
models, it is possible that the genes for enzymes involved in SCFA
synthesis may vary based on the type of colorectal tumor; where a
smaller number of microbes are responsible for the observed SCFA
concentrations in patients with carcinoma.

\textbf{Changes in genes for enzymes involved in SCFA synthesis are not
associated with tumors.} Using a list of specific genes that are
important for the synthesis of SCFAs {[}Table S1{]}, we looked for
differences in gene abundance based on individuals having normal colons,
adenomas, or carcinomas. First, using imputed gene relative abundance
based on 16S rRNA gene sequencing we found no difference in any of the
genes inolved with acetate, butyrate, or propionate synthesis (P-value
\textgreater{} 0.90) {[}Table S2{]}. This similarity between groups is
highlighted by visualizing genes important in butyrate synthesis
{[}Figure 3A{]}. Using a Wilcoxson rank-sum test, there also was no
difference in imputed gene relative abundance between pre- and
post-treatment samples for any genes involved with SCFA synthesis
(P-value \textgreater{} 0.70) {[}Table S3{]}. Next, we took a subset of
these 490 fecal samples (n=85) and used metagenomic sequencing to
confirm these results. Like the imputed gene results, metagenomic
analysis found that there was no difference in any of the genes involved
in SCFA synthesis between individuals with normal colons (n=29) and
patients with adenoma (n=28) or carcinoma (n=28) (P-value \textgreater{}
0.70) {[}Table S4{]}. This lack of difference is highlighted when we
visualize the results for butyrate kinase {[}Figure 3B{]}. These
observations provide evidence that the gene content does not change due
to colorectal tumors. Although these results suggest that SCFA
concentrations do not change due to colorectal tumor, there may be some
errors present within our results that may account for the lack of
associations.
